Hourly Assimilation of Different Sources of Observations Including Satellite Radiances in a  Mesoscale Convective System Case During RELAMPAGO campaign

### 2022-08-08 analysis/paper/paper_ana_20181122.Rmd
case overview, smothed conours and others

scale_date <- function(ini = 20181120180000, end = 20181123120000,
                       break_bin = "6 hour") {
  scale_x_datetime(breaks = seq.POSIXt(ymd_hms(ini), ymd_hms(end), by = break_bin),
                   labels = function(x) {
                     fifelse(hour(x) == 0, format(x, "%H UTC\n%b %d"), format(x, "%H"))
                   }) 
}

@@

For DA methods to be successful, observing networks with sufficient temporal and spatial resolution capable of capturing mesoscale variability should be used. In that regard, several authors investigated the impact of assimilating surface weather data (e.g. @wheatley2010, @ha2014, @chang2017, @bae2022, @banos2021, @maejima2019, and @chen2016) using different assimilation methodologies. Most of these studies reported the beneficial impacts of assimilating temperature and dew point observations upon the planetary boundary layer structure and the location and timing of precipitating systems.  @sobash2015 showed using a mesoscale DA system, that the positive impact upon convection initiation and the short range precipitation forecast is achieved if data is assimilated frequently (in the order of minutes, rather than in the order of hours).  More recently, @gasperoni2018 evaluated the impact of assimilating observations produced by private weather station networks which are not incorporated in the operational analysis. They found a positive effect of these observations upon the initiation of deep moist convection. This result is particularly important for data sparse regions such as Southern South America, where operational networks are not dense enough to capture mesoscale details.

@@

On Nov 22, 2018 a cold front crossed the center of Argentina triggering isolated convective cells (Figure \@ref(fig:case)a) that rapidly grew upscale into an exceptionally large MCS. During that day several surface stations reported lightning, gustfronts and heavy rain. To the north of the region, a warm front contributed to the development of isolated multicells that ultimately grew and merged with the MCS (Figure \@ref(fig:case)c). 
The MCS traveled approximately 2500 km from south to north, dissipating over Paraguay and Southern Brazil after 42 hours.

(ref:case) GOES-16 infrared imagery corresponding to channel 13 (13 mu) for a) 00 UTC, b) 12 UTC Nov 22, and c) 00 UTC Nov 23.

@@

campos <-  ReadNetCDF(here("analysis/data/derived_data/reanalysis/cape-pw.nc"), vars = "tcw")

@@

  .[time != ymd_h(2018112312)] %>% 
  .[, ":="(espesor = (`500`-`1000`)/100, time = factor(time))] %>%
  ggplot(aes(longitude, latitude)) +
  geom_contour_fill(aes(z = cape, fill = stat(level_d)), breaks = seq(500, 4000, 500)) +
  scale_fill_distiller(super = ScaleDiscretised,
                       name = "CAPE",
                       # palette = "PRGn", direction = 1,
                       palette = "YlOrRd", direction = 1,
                       guide = guide_colorsteps(barwidth = 25, mid = 25,
                                                barheight = 0.5, 
                                                title.position = "left", 
                                                title.vjust = 1,
                                                frame.colour = "black")) +
  geom_mapa() +
  geom_contour2(aes(z = espesor, label = ..level..),
                size = 0.2, color = "darkred", linetype = 2,
                label_size = 2, label.placer = label_placer_n(n = 3)) +
  # label_size = 2, label.placer = label_placer_fraction(frac = 0.75)) +
  geom_contour2(aes(z = msl/100, label = ..level..),
                size = 0.4, color = "black", linetype = 1,
                label_size = 2) +
  cordillera +
  coord_sf(ylim = c(-45, -22), xlim = c(-75, -52)) +
  facet_wrap(vars(time), labeller = labeller(time = c("2018-11-22 00:00:00" = "00 UTC Nov 22",
                                                      "2018-11-22 12:00:00" = "12 UTC Nov 22",
                                                      "2018-11-23 00:00:00" = "00 UTC Nov 23"))) +
  labs(x = NULL, y = NULL, fill = "CAPE" ) +
  theme_minimal() +
  theme(legend.position = "bottom")

@@

Figure \@ref(fig:rcrv-profile) shows the mean and standard deviation of the RCRV for the upper-air observations. Figures \@ref(fig:rcrv-profile)a-b show the RCRV statistics for soundings (ADPUPA) and aircraft (AIRCAR and AIRCFT). Both ADPUPA and AIRCFT show a generally good agreement between the ensemble spread and the observation error. As sounding observations and their associated errors are known to be reliable, this result indicates that the ensemble has an appropriate spread. AIRCAR presents an irregular profile with $sd RCRV$ values that suggest that the error for this type of observation is overestimated. ADPUPA and AIRCAR present a $mean RCRV$ profile near zero at middle and upper levels but ADPUPA observations show a cold bias with respect to the model at low levels, a characteristic already studied in @ruiz2010 and @dillon2021.

@@

ana[bottom_top %between% c(1, 7), .(v_mean = mean(v)), 
    by = .(south_north, west_east, date, exp)] %>% 
  .[pw, on = .NATURAL] %>% 
  ggplot(aes(x, y)) +
  geom_contour_fill(aes(z = value, fill = stat(level_d)),
                    proj = norargentina_lambert) +
  colorspace::scale_fill_continuous_divergingx(super = ScaleDiscretised,
                                               palette = "PRGn", mid = 25,
                                               guide = guide_colorsteps(barwidth = 25,
                                                                        barheight = 0.5, 
                                                                        title.position = "left", 
                                                                        title.vjust = 1,
                                                                        frame.colour = "black")) +
  geom_contour2(aes(z = v_mean, size = stat(level)),
                proj = norargentina_lambert, color = "black", 
                breaks = c(-10, -5)) +
  scale_size_continuous(breaks = c(-10, -5), range = c(0.8, 0.4),
                        labels = c("-10", "-5")) +
  labs(fill = latex2exp::TeX("Precipitable\nwater ($Kg$ $m^{-2}$)"), size = latex2exp::TeX("V ($m$ $s^{-1})")) +
  cordillera +
  geom_mapa() +
  facet_grid(. ~ exp, 
             labeller = labeller(exp = c(E2 = "CONV", E5 = "AWS", 
                                         E6 = "SATWND", E9 = "RAD"))) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.ontop = TRUE,
        panel.grid = element_line(linetype = 3)) +

  ana[bottom_top %between% c(1, 10), .(theta_mean = mean(theta)), 
      by = .(south_north, west_east, date, exp)] %>% 
  .[pw, on = .NATURAL] %>% 
  ggplot(aes(x, y)) +
  geom_contour_fill(aes(z = theta_mean, fill = stat(level_d)),
                    proj = norargentina_lambert,
                    breaks = c(-Inf, seq(290, 314, 2), Inf), limits = c(-Inf, Inf)) +
  scale_fill_distiller(super = ScaleDiscretised,
                       palette = "RdYlBu", direction = -1,
                       guide = guide_colorsteps(barwidth = 25,
                                                barheight = 0.5, 
                                                title.position = "left", 
                                                title.vjust = 1,
                                                frame.colour = "black")) +
  labs(fill = "Potential\ntemperature (K)") +
  cordillera +
  geom_mapa() +
  facet_grid(. ~ exp, 
             labeller = labeller(exp = c(E2 = "CONV", E5 = "AWS", 
                                         E6 = "SATWND", E9 = "RAD"))) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.ontop = TRUE,
        panel.grid = element_line(linetype = 3),
        strip.text = element_blank(), 
        strip.background = element_blank()) +

  mucape %>% 
  .[variable == "mcape"] %>% 
  .[, c("x", "y") := wrf_project(lon, lat)] %>% 
  .[cor, on = .NATURAL] %>% 
  ggplot(aes(x, y)) +
  geom_contour_fill(aes(z = value, fill = stat(level_d)), 
                    proj = norargentina_lambert, limits = c(500, 6000)) +
  scale_fill_distiller(super = ScaleDiscretised,
                       palette = "YlOrRd", direction = 1,
                       guide = guide_colorsteps(barwidth = 25,
                                                barheight = 0.5, 
                                                title.position = "left", 
                                                title.vjust = 1,
                                                frame.colour = "black")) +
  geom_contour2(aes(z = cortante, size = stat(level)),
                proj = norargentina_lambert, color = "black", 
                breaks = c(15, 30)) +
  scale_size_continuous(breaks = c(15, 30), range = c(0.8, 0.4),
                        labels = c("15", "30")) +
  labs(fill = latex2exp::TeX("MCAPE ($J$ $Kg^{-1}$)"), size = latex2exp::TeX("Wind\nshear ($m$ $s^{-1}$)")) +
  cordillera +
  geom_mapa() +
  facet_grid(. ~ exp, 
             labeller = labeller(exp = c(E2 = "CONV", E5 = "AWS", 
                                         E6 = "SATWND", E9 = "RAD"))) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.ontop = TRUE,
        panel.grid = element_line(linetype = 3),
        strip.text = element_blank(), 
        strip.background = element_blank()) +

@@

To complement the analysis, Figure \@ref(fig:dbz-mean) shows the observed maximum reflectivity in the vertical column (COLMAX) and the ensemble mean COLMAX for the CONV and RAD experiments at different times between 10 and 19 UTC Nov 22. These experiments were chosen because they represent the analysis with the minimum (CONV) and maximum (RAD) number of assimilated observations. In addition, they are the worst (CONV) and best (RAD) performing experiments in terms of the 1-hour precipitation forecast skill (Figure \@ref(fig:fss)). Overall, none of the short-range forecasts capture the mesoscale details in the reflectivity distribution. This is partially expected considering the coarse horizontal grid spacing (10 km), which is not enough to appropriately represent the strength of the convective band associated with the MCS. Also, the ensemble mean produces a smoothing effect that reduces the amplitude of local maxima in the forecast reflectivity field. RAD better represents the observed features of the system showing a  stronger and more organized MCS than CONV, over the domain center at 10 and 13 UTC (first and second columns in Figure \@ref(fig:dbz-mean)). The convective cells that initiate after 16 UTC along the warm front in the northeast part of the domain are well captured by both experiments but are better represented in terms of strength in RAD. In addition, CONV captures the location of the MCS, but the convection seems to be less organized and much weaker than in RAD. Before and after the times shown in Figure \@ref(fig:dbz-mean), the agreement between experiment and observations is quite good in the regions where radar data are available, especially for RAD.

@@

IOP 7 (Figures \@ref(fig:soundings)a-d) provides a good characterization of the pre-convective environment during the first day of our experiments. The area where the observations were taken was characterized by mostly clear skies and a low-level northerly flow associated with warm and moist advection. In general, the experiments show a similar RMSE and bias for all the variables. AWS observations were able to reduce the RMSE for temperature and dew point temperature in the PBL and reduce a small dry bias. However, AWS degradated the zonal wind between 7 and 12 km increasing the bias and RMSE (Figure \@ref(fig:soundings)c).

@@

In this paper, we investigate for the first time in South America the impact of different observation systems on the performance of an ensemble-based mesoscale regional DA system. In particular, we explore the impact upon the analysis quality of assimilating frequent and relatively dense surface observations, satellite derived winds, and satellite clear-sky radiances from multiple sensors. Southern South America is a particularly interesting region due to the heterogeneity and sometimes coarse resolution of the operational observing network (considering both surface based and upper air observations). This combined with a climatology characterized by frequent organized convective events makes mesoscale DA particularly challenging.  To consider these issues, we used a case study approach of a massive mesoscale convective system that developed over Southern South America, particularly in a region where the surface and upper-air observing networks are too coarse to capture variability at the mesoscale and that took place on Nov 22, 2018.

>>>>

scale_date <- function(ini = 20181120180000, end = 20181123120000,
                       break_bin = "6 hour", ...) {
  scale_x_datetime(breaks = seq.POSIXt(ymd_hms(ini), ymd_hms(end), by = break_bin), ..., expand = c(0, 0),
                   labels = function(x) {
                     fifelse(hour(x) == 0, format(x, "%H UTC\n%b %d"), format(x, "%H"))
                   }) 
}

@@

For DA methods to be successful, observing networks with sufficient temporal and spatial resolution capable of capturing mesoscale variability should be used (@gustafsson2018). In that regard, several authors investigated the impact of assimilating surface weather data (e.g. @wheatley2010, @ha2014, @chang2017, @bae2022, @banos2021, @maejima2019, and @chen2016) using different assimilation methodologies. Most of these studies reported the beneficial impacts of assimilating temperature and dew point observations upon the planetary boundary layer structure and the location and timing of precipitating systems.  @sobash2015 showed using a mesoscale DA system, that the positive impact upon convection initiation and the short range precipitation forecast is achieved if data is assimilated frequently (in the order of minutes, rather than in the order of hours).  More recently, @gasperoni2018 evaluated the impact of assimilating observations produced by private weather station networks which are not incorporated in the operational analysis. They found a positive effect of these observations upon the initiation of deep moist convection. This result is particularly important for data sparse regions such as Southern South America, where operational networks are not dense enough to capture mesoscale details.

@@

Previoursly to the development of this case study, the center and north of Argentina was inmersed in a warm and humid air mass with high values of convective available potential energy (CAPE) (Figure \@ref(fig:case)a). On Nov 22, 2018 a cold front crossed the center of Argentina triggering isolated convective cells (Figure \@ref(fig:case)b) that rapidly grew upscale into an exceptionally large MCS. During that day several surface stations reported lightning, gustfronts and heavy rain. To the north of the region, a warm front contributed to the development of isolated multicells that ultimately grew and merged with the MCS (Figure \@ref(fig:case)c). 
The MCS traveled approximately 2500 km from south to north, dissipating over Paraguay and Southern Brazil after 42 hours.

(ref:case) ERA5 Reanalysis of pressure (hPa) at sea level (black contours), 1000-500 hPa thickness (red dashed contours) and convective available potential energy (shaded) for 00 and 12 UTC Nov 22 and 00 UTC Nov 23.

@@

# campos <-  ReadNetCDF(here("analysis/data/derived_data/reanalysis/cape-pw.nc"), vars = "tcw")

@@

  .[time != ymd_h(2018112312)] %>% 
  .[, ":="(espesor = (`500`-`1000`)/100, time = factor(time))] %>%
  ggplot(aes(longitude, latitude)) +
  geom_contour_fill(aes(z = cape, fill = stat(level_d)), breaks = seq(500, 4000, 500)) +
  scale_fill_distiller(super = ScaleDiscretised,
                       name = "CAPE",
                       # palette = "PRGn", direction = 1,
                       palette = "YlOrRd", direction = 1,
                       guide = guide_colorsteps(barwidth = 25, mid = 25,
                                                barheight = 0.5,
                                                title.position = "left",
                                                title.vjust = 1,
                                                frame.colour = "black")) +
  geom_mapa() +
  geom_contour2(aes(z = espesor, label = ..level..),
                size = 0.2, color = "darkred", linetype = 2,
                label_size = 2, label.placer = label_placer_n(n = 3)) +
  # label_size = 2, label.placer = label_placer_fraction(frac = 0.75)) +
  geom_contour2(aes(z = msl/100, label = ..level..),
                size = 0.4, color = "black", linetype = 1,
                label_size = 2) +
  cordillera +
  coord_sf(ylim = c(-45, -22), xlim = c(-75, -52)) +
  facet_wrap(vars(time), labeller = labeller(time = c("2018-11-22 00:00:00" = "00 UTC Nov 22",
                                                      "2018-11-22 12:00:00" = "12 UTC Nov 22",
                                                      "2018-11-23 00:00:00" = "00 UTC Nov 23"))) +
  tag_facets() +
  labs(x = NULL, y = NULL, fill = "CAPE" ) +
  theme_minimal() +
  theme(legend.position = "bottom")

@@

Figure \@ref(fig:rcrv-profile) shows the mean and standard deviation of the RCRV for the upper-air observations. Figures \@ref(fig:rcrv-profile)a-b show the RCRV statistics for soundings (ADPUPA) and aircraft (AIRCAR and AIRCFT). Both ADPUPA and AIRCFT show a generally good agreement between the ensemble spread and the observation error. As sounding observations and their associated errors are known to be reliable, this result indicates that the ensemble has an appropriate spread. AIRCAR presents an irregular profile with $sd RCRV$ values that suggest that the error for this type of observation is overestimated. ADPUPA and AIRCAR present a $mean RCRV$ profile near zero at middle and upper levels. At low levels, the $mean RCRV$ profile is positive, showing a cold bias present in the model, a characteristic already studied in @ruiz2010 and @dillon2021.

@@

ana[bottom_top %between% c(1, 7), .(v_mean = mean(v)), 
    by = .(south_north, west_east, date, exp)] %>% 
  .[pw, on = .NATURAL] %>% 
  .[, v_mean_suave := metR:::smooth2d(x, y, v_mean, kx = 0.2, ky = 0.2), by = .(exp)] %>% 
  ggplot(aes(x, y)) +
  geom_contour_fill(aes(z = value, fill = stat(level_d)),
                    proj = norargentina_lambert) +
  colorspace::scale_fill_continuous_divergingx(super = ScaleDiscretised,
                                               palette = "PRGn", mid = 25,
                                               guide = guide_colorsteps(barwidth = 25,
                                                                        barheight = 0.5, 
                                                                        title.position = "left", 
                                                                        title.vjust = 1,
                                                                        frame.colour = "black")) +
  geom_contour2(data = ~.[x %between% c(-950000, 950000)], aes(z = v_mean_suave, size = stat(level)),
                proj = norargentina_lambert, color = "black", 
                breaks = c(-10, -5)) +
  scale_size_continuous(breaks = c(-10, -5), range = c(0.6, 0.4),
                        labels = c("-10", "-5")) +
  labs(fill = latex2exp::TeX("Precipitable\nwater ($Kg$ $m^{-2}$)"), size = latex2exp::TeX("V ($m$ $s^{-1})")) +
  cordillera +
  geom_mapa() +
  facet_grid(. ~ exp, 
             labeller = labeller(exp = c(E2 = "CONV", E5 = "AWS", 
                                         E6 = "SATWND", E9 = "RAD"))) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.ontop = TRUE,
        panel.grid = element_line(linetype = 3)) +

  ana[bottom_top %between% c(1, 10), .(theta_mean = mean(theta)), 
      by = .(south_north, west_east, date, exp)] %>% 
  .[pw, on = .NATURAL] %>% 
  ggplot(aes(x, y)) +
  geom_contour_fill(aes(z = theta_mean, fill = stat(level_d)),
                    proj = norargentina_lambert,
                    breaks = c(-Inf, seq(290, 314, 2), Inf), limits = c(-Inf, Inf)) +
  scale_fill_distiller(super = ScaleDiscretised,
                       palette = "RdYlBu", direction = -1,
                       guide = guide_colorsteps(barwidth = 25,
                                                barheight = 0.5, 
                                                title.position = "left", 
                                                title.vjust = 1,
                                                frame.colour = "black")) +
  labs(fill = "Potential\ntemperature (K)") +
  cordillera +
  geom_mapa() +
  facet_grid(. ~ exp, 
             labeller = labeller(exp = c(E2 = "CONV", E5 = "AWS", 
                                         E6 = "SATWND", E9 = "RAD"))) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.ontop = TRUE,
        panel.grid = element_line(linetype = 3),
        strip.text = element_blank(), 
        strip.background = element_blank())

  mucape %>% 
  .[variable == "mcape"] %>% 
  .[, c("x", "y") := wrf_project(lon, lat)] %>% 
  .[cor, on = .NATURAL] %>% 
  .[, cortante_suave := metR:::smooth2d(x, y, cortante, kx = 0.2, ky = 0.2), by = .(exp)] %>% 
  ggplot(aes(x, y)) +
  geom_contour_fill(aes(z = value, fill = stat(level_d)), 
                    proj = norargentina_lambert, limits = c(500, 6000)) +
  scale_fill_distiller(super = ScaleDiscretised,
                       palette = "YlOrRd", direction = 1,
                       guide = guide_colorsteps(barwidth = 25,
                                                barheight = 0.5, 
                                                title.position = "left", 
                                                title.vjust = 1,
                                                frame.colour = "black")) +
  geom_contour2(data = ~.[x %between% c(-950000, 950000) &
                           y %between% c(-1170000, 1170000)], aes(z = cortante_suave, size = stat(level)),
                proj = norargentina_lambert, color = "black", 
                breaks = c(15, 30)) +
  scale_size_continuous(breaks = c(15, 30), range = c(0.6, 0.4),
                        labels = c("15", "30")) +
  labs(fill = latex2exp::TeX("MCAPE ($J$ $Kg^{-1}$)"), size = latex2exp::TeX("Wind\nshear ($m$ $s^{-1}$)")) +
  cordillera +
  geom_mapa() +
  facet_grid(. ~ exp, 
             labeller = labeller(exp = c(E2 = "CONV", E5 = "AWS", 
                                         E6 = "SATWND", E9 = "RAD"))) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.ontop = TRUE,
        panel.grid = element_line(linetype = 3),
        strip.text = element_blank(), 
        strip.background = element_blank()) +

@@

To complement the analysis, Figure \@ref(fig:dbz-mean) shows the observed maximum reflectivity in the vertical column (COLMAX) and the ensemble mean COLMAX for the CONV and RAD experiments at different times between 10 and 19 UTC Nov 22. These experiments were chosen because they represent the analysis with the minimum (CONV) and maximum (RAD) number of assimilated observations. In addition, they are the worst (CONV) and best (RAD) performing experiments in terms of the 1-hour precipitation forecast skill (Figure \@ref(fig:fss)). Overall, none of the short-range forecasts capture the mesoscale details in the reflectivity distribution. This is partially expected considering the coarse horizontal grid spacing (10 km), which is not enough to appropriately represent the strength of the convective band associated with the MCS. Also, the ensemble mean produces a smoothing effect that reduces the amplitude of local maxima in the forecast reflectivity field. RAD better represents the observed features of the system showing a  stronger and more organized MCS than CONV, over the domain center at 10 and 13 UTC (first and second columns in Figure \@ref(fig:dbz-mean)). The convective cells that initiate after 16 UTC along the warm front in the northeast part of the domain are well captured by both experiments but are better represented in terms of strength in RAD. In addition, CONV captures the location of the MCS, but the convection seems to be less organized and much weaker than in RAD. Before and after the times shown in Figure \@ref(fig:dbz-mean), the agreement between location of the observed convective cells and the simulated in the experiment is quite good in the regions where radar data are available, especially for RAD.

@@

IOP 7 (Figures \@ref(fig:soundings)a-d) provides a good characterization of the pre-convective environment during the first day of our experiments. The area where the observations were taken was characterized by mostly clear skies and a low-level northerly flow associated with warm and moist advection. In general, the experiments show a similar RMSE and bias for all the variables. AWS observations were able to reduce the RMSE for temperature and dew point temperature in the PBL and reduce a small dry bias. However, AWS increments (Figure \@red(fig:UV-diff)d) degradated the zonal wind between 7 and 12 km increasing the bias and RMSE (Figure \@ref(fig:soundings)c).

@@

In this paper, we investigate for the first time in South America the impact of different observation systems on the performance of an ensemble-based mesoscale regional DA system. In particular, we explore the impact upon the analysis quality of assimilating frequent and relatively dense surface observations, satellite derived winds, and satellite clear-sky radiances from multiple sensors. Southern South America is a particularly interesting region due to the heterogeneity in topography and weather and sometimes coarse resolution of the operational observing network (considering both surface based and upper air observations). This combined with a climatology characterized by frequent organized convective events makes mesoscale DA particularly challenging.  To consider these issues, we used a case study approach of a massive mesoscale convective system that developed over Southern South America, particularly in a region where the surface and upper-air observing networks are too coarse to capture variability at the mesoscale and that took place on Nov 22, 2018.
